The bhsdtr package: a general purpose method of Bayesian inference for Signal Detection Theory models

### 2018-02-13 inst/preprint/paper.tex
apa formatted paper

% WĄTPLIWOŚCI JĘZYKOWE oznaczone przez %?? !! oznacza miejsce, gdzie
% skończyłem ostatnio poprawiać

% przykłady w sieci value of <> [directly] represents the <>

% jak się pisze 16ms czy 16 ms

% argumenty funkcji są an, gdy piszemy ogólnie o działaniu tych
% funkcji

\documentclass[oneside,a4paper]{article}
\usepackage[sectionbib,bibnewpage]{apacite}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{float}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{caption}

\title{The bhsdtr package: a general purpose method of bayesian
  inference for Signal Detection Theory models}

\begin{document}
\maketitle
\tableofcontents{}

\section{Abstract}

We describe a novel method of bayesian inference for hierarchical or
non-hierarchical equal variance normal Signal Detection Theory models
with one or more criteria. The method is implemented as an open-source
R package that uses the state-of-the-art platform Stan for sampling
from posterior distributions. Our method can accommodate binary
responses as well as additional ratings and an arbitrary number of
nested
% ?? WSZYSTKIE sdt parameters CAN be regressed, więc the
or crossed random grouping factors. The SDT parameters can be
regressed on additional predictors within the same model via
intermediate unconstrained parameters, and the model can be extended
by using automatically generated human-readable Stan code as a
% ?? czy to jest przypadek other -> bez przedimka?
template. In the paper we explain how our method improves on other
similar available methods, we give an overview of the package,
demonstrate the ease of use by providing a real-study data analysis
walk-through, and show that the model successfully recovers known
parameter values when fitted to simulated data.

\section{Introduction}

Many tasks used in psychology studies are essentially classification
tasks. For example, in a memory study participants may be required to
decide if a given test item is old or new, or, in a perceptual study,
an object may either be a letter or a digit. If a task requires
classification it is always possible that conclusions based on
% ? an ability mi tu nie pasuje
accuracy or percent correct are invalid because ability to
discriminate between stimulus classes (i.e., sensitivity) is
confounded with bias, which is a tendency to classify stimuli as
belonging to a particular class \cite{GreenSwets66}. In principle, any
effect that manifests itself in differences in classification accuracy
% ?? ani the sensitivity ani a sensitivity nie pasuje. Chodzi o
% ogólnie rozumiane zjawisko.
may reflect differences in sensitivity, bias, or both. Signal
Detection Theory provides a simple and popular solution to this common
and important problem.

However, because the SDT model is non-linear any variability in its
parameters due to factors such as participants or items has to be
accounted for, otherwise the estimates of SDT parameters are
biased. As we explain later in this paper, neither of the two
available methods of inference for hierarchical SDT models that we are
aware of addresses this problem correctly. Hence, our main goal was to
create a correct implementation of the general hierarchical linear
regression structure defined on SDT parameters. We argue that the
\texttt{bhsdtr} package for R \cite{rstatistical} provides exactly
such an implementation and we have made it publicly available at
% ?? tylko jeden, konkretny kod i zbiór danych, o którym czytelnik
% dopiero się dowiaduje, ale o którym nie będziemy więcej mówić, czyli
% the
https://github.com/boryspaulewicz/bhsdtr, together with the annotated
source code that was used to perform all the analyses and produce all
the figures presented in this paper.

% OPIS STRUKTURY ARTYKUŁU

In what follows, after introducing the most common version of the SDT
model, we describe its generalization which can accommodate data from
rating experiments. Next, we explain briefly why, if a method of
inference for SDT models were to be of general use in psychology
studies, it is essential that it was based on a model equipped with
the general hierarchical linear regression structure. The
\texttt{bhsdtr} package meets this requirement thanks to a novel
parametrization. We describe this novel parametrization in great
detail and we motivate it by explaining how the reliance on a
relatively standard parametrisation leads to problems in the two other
available implementations. We end the first part of this paper with a
formal definition of the model as implemented in \texttt{bhsdtr}. The
second part contains an overview of the package and a tutorial in
which we demonstrate how to use our method in practice. Before we go
any further, however, a note on terminology seems in order.

% SAMPLED FACTOR

In the context of hierarchical modelling factors such as participants,
items, or replications are often referred to as groups. In our opinion
this naming convention may be confusing; a single participant is both
a group and a member of some group, while at the same time the term
"group" is perhaps most strongly associated with study conditions, as
in "experimental group". In this paper we use the term "sampled
factor" instead, because, by virtue of being a new term, it is
unambiguous, and because it seems descriptively correct: the term
"sampled factor" captures all the essential properties of such
variables, i.e., a nominal scale, the fact that values are sampled
from a larger population and are usually not of direct interest, as in
"this is only a sample", and that conclusions
% ?? jeżeli już mamy jakąś analizę statystyczną, to zakładamy, w
% uproszczeniu, że target population jest określona
of statistical analysis are meant to apply to the whole population of
possible values.

\section{Equal variance normal Signal Detection Theory model with
  additional criteria}

According to Signal Detection Theory each stimulus in a classification
task gives rise, by some unspecified cognitive process, to an
unidimensional internal evidence value $s$ sampled from
% ?? stimulus już był wprowadzony, więc jego klasa jest (jakaś, ale)
% określona
a distribution that depends on the stimulus class. For historical
reasons the two stimulus classes are often referred to as "noise" and
"signal", and task performance is described in terms of hits, correct
rejections, omissions, and false alarms, but this terminology is
appropriate only when the model is applied to tasks that require
detection, which is by far not always the case. In the most widely
used version of the model, shown in Fig.~\ref{basicsdt}, the two
evidence distributions are normal with the same variance, which is
usually fixed at unity to make the model identifiable. The distance
$d'$ between the means of the
% ?? sensitivity bez przedimka?
evidence distributions represents sensitivity. Because normal
distributions are unbounded, $s$ is always ambiguous, and so a
criterion $c$ placed on the evidence axis has to be used to reach a
discrete
% ?? bias bez przedimka?
decision. A participant is assumed to decide that a stimulus belongs
to the first class (e.g., an old item) if $s < c$, or that it belongs
to the second class (e.g., a new item) if $s \geq c$. The location of
the decision criterion $c$ represents bias.

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot_basicsdt.pdf}
  \caption{Equal variance normal Signal Detection Theory model}
  \label{basicsdt}
\end{figure}

% Najprostsze użycie

Perhaps the simplest way of using this model is to fit it to observed
response counts and use the estimated $d'$ values in place of the
percent correct scores; If the model is correct, the resulting
performance measure is not contaminated by bias. Obviously, the model
may not be correct, which is one of the reasons why we focus more on
the generalized version shown in Fig.~\ref{sdtr} below.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot_sdtr.pdf}
  \caption{Equal variance normal Signal Detection Theory model with
    additional criteria}
  \label{sdtr}
\end{figure}

This
% ?? applicable to czy applicable in
generalized model is applicable to studies where participants are
asked to rate their binary classification decisions on confidence or
some other performance or stimulus related
% the ratings, bo study już określone, podobnie jak (rodzaj)
% klasyfikacji binarnej, ale czy the ratings and THE binary ...?
dimension. The ratings and the binary classification decisions can be
provided together (e.g., "I am almost certain that it was a digit"),
or in an arbitrary order.

% MODELOWANIE RATINGÓW ZA POMOCĄ KRYTERIÓW

% ?? a combined response? Skoro mamy ratings (i w domyśle mamy też
% binary class.), to czy nie the combined response? A może
% jakiekolwiek ratingi są załatwione za pomocą (ogólnego rozwiązania)
% modelowania czegoś, co jest "odpowiedzią łączną"?
Ratings are accommodated by introducing additional criteria and
modelling a combined response $y$, which
% ?? znowu kwestia przedimków vs. and
represents both the binary classification decision and rating. The
value of $y$ increases with the strength of evidence in favor of the
second stimulus class. For example, if confidence is rated on a
four-point scale, then $y = 1$ when a subject decides
% ?? bodziec, który zobaczył badany, więc the? bo chodzi o decyzję o
% określonym dla decydenta bodźcu, a nie nieokreślonym dla decydenta?
that a stimulus belongs to the first class with certainty $4$, and
$y = 5$ when a subject decides that a stimulus belongs to the second
class with certainty $1$. More formally, if $K$ is the number of
possible combined responses then a subject is assumed to give response
$k$ if $s \in (\bm{c}_{k-1},\bm{c}_k]$, where $\bm{c}$ are the $K+1$
decision criteria, with $\bm{c}_0$ and $\bm{c}_K$ fixed at $-\infty$
and $+\infty$ respectively.

% UŻYWAĆ SDTR CHOĆBY PO TO, ŻEBY SPRAWDZIĆ JEGO ZAŁOŻENIA

There is a good reason to collect the ratings and use the generalized
SDT model from Fig.~\ref{sdtr} even when neither the ratings nor the
placement of criteria are relevant to the research problem.
% ?? If K = 2 to jakby: w (jakimś) przypadku, w którym, When K = 2 to
% jakby: zawsze, gdy ...? Czuję, że lepiej the data i the model potem
When $K = 2$ (no ratings), the SDT model fits perfectly, regardless of
it being a reasonably good approximation to reality or not, because
the data and the model have the same dimensionality. This makes the
generalization to the $K > 2$ case particularly important, as it is
% ?? niektóre lub wszystkie formal assumptions, ale mamy na myśli
% konkretny zbiór formal assumptions, który czytelnik już zna, a
% nawet, jeśli nie, to te założenia są unikalnie wyznaczone przed
% model - the?
only when $K > 2$, that the formal assumptions of the model (e.g.,
equal or unequal variance) can be tested\footnote{In contrast to the
  formal assumptions, a psychological interpretation of the SDT
  parameters can be tested even when ratings are not available, e.g.,
  by means of selective influence \cite{sternberg2001separate}}, which
is often done by comparing observed and predicted ROC curves.

When SDT models are used in psychology studies, researchers are
usually interested in relationships between SDT parameters and
additional measured or manipulated
% ?? the dependence, bo ogólnie rozumiana kwestia zależności, stimulus
% strength - ogólne zjawisko, a nie jakaś konkretna, albo pojedyncza,
% ale bliżej nieokreślona
variables; A good example is the dependence of $d'$ on stimulus
strength. Also, in a great majority of psychology studies in which
classification tasks are used the data have hierarchical structure,
i.e., there are repeated measures for each participant or item and
participants or items are only samples from the target population. It
seems worthwhile to explain why a general-purpose method of inference
for SDT models should accommodate both kinds of situations.

% ?? Ważność hierarchical structure jako takiej
\subsection{The importance of hierarchical regression structure}

If data have hierarchical structure, but variability due to subjects,
items, or other sampled factors is not accounted for, estimates of
average (fixed) effects are not guaranteed to be unbiased and
conclusions are not guaranteed to generalize to a target population. A
not so uncommon practice of analysing data aggregated over sampled
factors represents an extreme case of ignoring
% ?? nieokreślona hier. structure, bo tu mówimy o externe case
% ogólnego zjawiska ignorowania ogólnej kwestii hierarchiczności
hierarchical data structure. Invalidity of this approach in the
% ?? Jeżeli illustrated w czasie przeszłym dokonanym i mówimy zaraz
% czyje rezultaty, to chyba the results?
context of SDT was clearly illustrated with the results of
simulational studies by \citeA{morey2008problematic}, although,
strictly speaking, it requires such demonstrations only in order to
make a point about a specific case. Because SDT is a non-linear model,
% ?? data jest unikalne dla tokenów estimates, więc the data?
by definition, when estimates of its parameters are based on data
aggregated over sampled factors the expected values of these estimates
are not in general true population averages. Such estimates are
% ?? zakładamy (w uproszczeniu), że target population jest ustalona
% dla każdego estimate
biased and inference about a target population based on such estimates
is simply not valid. To give a concrete example, consider two unbiased
participants, one with $d'_1 = 2$ and one with $d'_2 = 4$. Their
expected average accuracy is given by
$(\Phi(d'_1/2) + \Phi(d'_2/2)) / 2 = .91$, which corresponds to
$d' = 2.68$, whereas their true average $d'$ is $3$.

Aggregation is not the only way of ignoring hierarchical data
structure. Sometimes non-aggregated data are analyzed by using
separate estimates for every participant $\times$ item $\times$
% ?? mamy już dane i estimates, to (zwykle nieznany) (poziom)
% uncertainty jest określony
condition combination, but the uncertainty due to subject or item
effects \emph{distributions} is not accounted for by means of a
hierarchical model
% dalej jedziemy z nieokreślonymi, bo wyjaśniamy ogólne zasady:
% wnioski mogą być jakiekolwiek, próbki i populacje są bliżej
% nieokreślone
structure. In such cases conclusions, at least with respect to
% ?? może chodzić o różne population-average różnie estymowalne
%
% ?? population-average - or fixed - effects? chodzi mi o to, że
% effects ma się też "łączyć" z population-average
the uncertainties in estimates of population-average (fixed)
% ?? sample i population jest dana dla tokenów
effects, are guaranteed to be valid only for the given sample, not the
target population.

% mówimy o ogólnym co prawda, ale zarazem pojedynczym/unikalnym
% problemie: the separation? Chętnie bym się pozbył przedimka
Another common source of problems is the separation of estimation of
non-linear model parameters from regression analysis which aim to
relate these parameters to measured or manipulated variables. When the
SDT parameters are estimated separately for each subject and
condition, and only later regressed on predictors of interest, a
number of additional issues may arise.

% tutaj jakaś regression analysis już jest wcześniej wprowadzona
Firstly, the standard errors or credible intervals associated with the
regression coefficients do not reflect the uncertainty in the SDT
parameter estimates, because the latter are treated as mere data
points. Precision of parameter estimates often varies between
participants, items, or conditions, but when the estimates are treated
as data points no use is made of this information. Secondly,
regressing parameters on numeric predictors makes their estimates
dependent on the common regression structure, and so also on each
other, which can improve the quality of the estimates, just as
assuming that random effects are samples from a common distribution
may improve their estimates.

The aforementioned problems can be dealt with by supplementing the SDT
model with the general hierarchical linear regression structure. For
the reasons that we will now explain, this can only be achieved if the
model is substantially reparametrised.

% dopiero wprowadzamy ową unconstrained parameter space
\section{Hierarchical Signal Detection Theory in a constrained
  parameter space}

Both $d'$ and $c$ have the virtue of being directly interpretable in
terms of sensitivity and bias. However, both $d'$ and $c$ are
constrained: $d'$ is non-negative and, when there is more than one
criterion, the elements of the $\bm{c}$ vector are order-restricted
($\bm{c}_{i+1} > \bm{c}_i$). Because normal distribution is unbounded,
hierarchical linear regression structure, in which random effects are
normally distributed and fixed effects are represented by
unconstrained parameters, can only be defined on unconstrained
parameters. We provide examples of the problems that may otherwise
arise in the following summary of hierarchical SDT implementations.

We are aware of only two attempts at creating a general purpose method
of inference for hierarchical SDT models with ratings. One is the
Gibbs sampler proposed by \citeA{morey2008problematic} and the other
is the Hierarchical Meta-d' model (HMeta-d') proposed by
\citeA{hmetad}.
% We do not consider the ROC Toolbox \cite{Koen2016} or other similar
% tools here because, as of this writing, these tools cannot be used to
% fit hierarchical SDT models. The two available methods that can be
% used to fit hierarchical SDT models are
The HMeta-d' model is a hierarchical version of the meta-d' model
\cite{maniscalco2012signal}, which in turn is a generalization of the
SDT model that allows for a separate "meta-sensitivity" to account for
possible discrepancies
% referred to as ... bez przyimka, prawda?
between a binary stimulus classification (referred to as type 1 task)
% ?? wcześniej było a binary i a rating, jakby to były (potencjalnie)
% niezwiązane zadania
and the associated rating task (referred to as type 2 or
meta-cognitive task). We consider HMeta-d' here because it reduces to
the SDT model with
% ?? powtórzenie the
ratings when the type 1 and the type 2 sensitivities are equal.

The Gibbs sampler created by \citeA{morey2008problematic} allows for
at most two sampled factors to have independent normally distributed
random effects on the evidence distribution means. Unlike $d'$, each
evidence distribution mean considered in isolation is an unconstrained
parameter, but the mean of the second one is by definition greater
($d' > 0$), or equal ($d' = 0$), than the mean of the first one. The
authors explicitly admit that their algorithm does not enforce this
restriction, because, as they claim in the paper, it would greatly
complicate analysis. At the same time the authors fail to mention that
an immediate consequence of allowing for negative $d'$ values is that
the resulting posterior distribution looses its intended
interpretation, since it can have a non-zero mass when $d' < 0$. The
outermost criteria are fixed at $0$ and $1$, and the ordering
restriction is enforced by assuming that the likelihood is $0$
whenever $\bm{c}_{i+1} \leq \bm{c}_i$. As the authors explain, because
a sampled factor can have independent random effects on the evidence
distribution means, it can have an effect on all the criteria:
shifting both means by the same amount in the same direction is
equivalent to keeping the sensitivity intact, while shifting the
criteria relative to the evidence distributions. However, the elements
of the criteria vector cannot be affected differently by the same
sampled factor, which is an unrealistic restriction.

In HMeta-d' the hierarchical structure is restricted to normally
distributed random intercepts of one sampled factor. Perhaps the most
problematic aspect of the HMeta-d' model, apart from the fact that
also in this model the $d'$ parameter is allowed to assume negative
values, is the representation of the criteria. Each element of the
criteria vector has an associated independent normal distribution,
which does allow for criteria random effects, but does not enforce the
necessary ordering restriction. The elements of this vector are sorted
to obtain another criteria vector, and it is this sorted criteria
vector that is used to model the conditional combined response
distributions. Consequently, the model does contain parameters
representing the actual, order-restricted criteria, but, because
sorting is not injective, the
% ?? (i.e., not isomorphic to)
space of the actual criteria is only loosely related (i.e., not
isomorphic) to the space of the unrestricted criteria vectors that are
associated with a hierarchical structure.

% Hierarchical structure does not have to be linear, but when it is, the
% parameters associated with unconstrained fixed effects and normally
% distributed random effects of sampled factors cannot be non-negative
% or order-restricted. We propose to substantially reparametrise the SDT
% model.

@@

The general hierarchical linear regression structure can be defined on
SDT parameters only if the latter are derived from unconstrained
parameters. In the \texttt{bhsdtr} package $d'$ is derived from
$\delta = \ln(d')$, this way random effects on $d'$ can be modelled by
assuming that $\delta$ is normally distributed. The problem of
representing the criteria by unconstrained parameters
% ?? skoro the space of, to potem the... ? w końcu chodzi nam o
% *wszystkie*, takie, że...?
is solved by mapping the $R^{K-1}$ space of unconstrained criteria
vectors to the $K$ dimensional probability simplex space using the
softmax function, and mapping the simplex space to the space of
order-restricted criteria vectors by means of the inverse normal CDF:

@@

\noindent Here $i = 1\dots N$ is the observation number, $\bm{X}$ is
the fixed effects model matrix for the respective parameter, $\bm{Z}$
is the random effects model matrix, $\bm{\beta}$ and $\bm{\theta}$ are
the
% opisać raczej jako macierz c, itd. - jest macierzą != jest efektem
% losowym
fixed and random effects, $\bm{c}$ is a $N \times K-1$ matrix, $s$ is
the criteria scaling factor, and $\bm{y}$ is the combined
response. Note that $\bm{d'}_i$ is a scalar, but
$\bm{\gamma}_{i,\cdot}$ is in general a vector, and so
$\bm{\beta}^{(\gamma)}$ and $\bm{\theta}^{(\gamma)}$ are matrices. The
$j$-th rows of the $\bm{\beta}^{(\gamma)}$ and
$\bm{\theta}^{(\gamma)}$ matrices
% ?? effects on ... ?
represent fixed and random effects on the $j$-th element of the
$\bm{\gamma}$ vector.

% ?? tutaj coś dziwnego dzieje się z liczbą: ta dekompozycja tych
% macierzy, i ta (jedna) dekompozycja poprawia próbkowanie itd.
Following \citeA{sorensen2015bayesian} we make use of the Cholesky
decomposition of the correlation matrices, because it improves
efficiency and admits a convenient prior on random effects
correlations:

>>>>

% -*- coding: utf-8 -*-

\documentclass[a4paper,man,apacite,floatsintext]{apa6}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{float}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{caption}

\title{The bhsdtr package: a general purpose method of Bayesian
  inference for Signal Detection Theory models}

\shorttitle{The bhsdtr package}

\abstract{We describe a novel method of Bayesian inference for
  hierarchical or non-hierarchical equal-variance normal Signal
  Detection Theory models with one or more criteria. The method is
  implemented as an open-source R package that uses the
  state-of-the-art platform, Stan, for sampling from posterior
  distributions. Our method can accommodate binary responses as well
  as additional ratings and an arbitrary number of nested or crossed
  random grouping factors. The SDT parameters can be regressed on
  additional predictors within the same model via intermediate
  unconstrained parameters, and the model can be extended by using
  automatically generated human-readable Stan code as a template. In
  the paper we explain how our method improves on other similar
  available methods, we give an overview of the package, demonstrate
  its ease of use by providing a real-study data analysis
  walk-through, and show that the model successfully recovers known
  parameter values when fitted to simulated data.}

\keywords{Signal Detection Theory, Bayesian inference, hierarchical
  models}

\twoauthors{Borysław Paulewicz}{Agata Blaut}

\twoaffiliations{University of Social Sciences and Humanities, Faculty
  in Katowice, Poland}{Department of Psychology, Jagiellonian
  University, Krakow, Poland}

\authornote{This research was supported in part by the National
  Science Centre, Poland HARMONIA grant given to Michal Wierzchon
  (2014/14/M/HS6/00911).

  Correspondence concerning this article should be addressed to
  Borysław Paulewicz, Department of Psychology, University of Social
  Sciences and Humanities, Faculty in Katowice, ul. Techników 9,
  40-326 Katowice, Poland.

Contact: bpaulewicz@swps.edu.pl}

\newcommand{\code}[1]{\texttt{#1}}

\begin{document}
\maketitle
Many tasks used in psychology studies are essentially classification
tasks. For example, in a memory study participants may be required to
decide if a given test item is old or new, or, in a perceptual study,
an object may be either a letter or a digit. If a task requires
classification, it is always possible that conclusions based on
accuracy or percent correct are invalid because the ability to
discriminate between stimulus classes (i.e., sensitivity) is
confounded with bias, which is a tendency to classify stimuli as
belonging to a particular class \cite{GreenSwets66}. In principle, any
effect that manifests itself in differences in classification accuracy
may reflect differences in sensitivity, bias, or both. Signal
Detection Theory provides a simple and popular solution to this common
and important problem.

However, because the SDT model is non-linear, variability in its
parameters due to factors such as participants or items has to be
accounted for, otherwise the estimates of SDT parameters are
biased. As we explain later in this paper, neither of the two
available methods of inference for hierarchical SDT models that we are
aware of addresses this problem correctly. Hence, our main goal was to
create a correct implementation of the general hierarchical linear
regression structure defined on SDT parameters. We argue that the
\code{bhsdtr} package for R \cite{rstatistical} provides exactly such
an implementation and we have made it publicly available at
\code{https://github.com/boryspaulewicz/bhsdtr}, together with the
annotated source code that was used to perform all the analyses and
produce all the figures presented in this paper.

In what follows, after introducing the most common version of the SDT
model, we describe its generalization, which can accommodate data from
rating experiments. Next, we explain briefly why, if a method of
inference for SDT models were to be of general use in psychology
studies, it is essential that it is based on a model equipped with the
general hierarchical linear regression structure. The \code{bhsdtr}
package meets this requirement thanks to a novel parametrization. We
describe this novel parametrization in great detail and explain how
reliance on relatively standard parametrization leads to problems in
the two other available implementations. We end the first part of this
paper with a formal definition of the model as implemented in
\code{bhsdtr}. The second part contains an overview of the package and
a tutorial in which we demonstrate how to use our method in
practice. Before we go any further, however, a note on terminology
seems in order.

In the context of hierarchical modelling, factors such as
participants, items, or replications are often referred to as
groups. In our opinion this naming convention may be confusing; a
single participant is both a group and a member of some group, while
at the same time the term "group" is perhaps most strongly associated
with study conditions, as in "experimental group". In this paper we
use the term "sampled factor" instead because, by virtue of being a
new term, it is unambiguous and seems descriptively correct: the term
"sampled factor" captures all the essential properties of such
variables, i.e., a nominal scale, the fact that values are sampled
from a larger population and are usually not of direct interest, as in
"this is only a sample", and that conclusions of statistical analysis
are meant to apply to the whole population of possible values.

\section{Equal-variance normal SDT model with additional criteria}

According to SDT, each stimulus in a classification task gives rise,
by some unspecified cognitive process, to a unidimensional internal
evidence value $s$ sampled from a distribution that depends on the
stimulus class. For historical reasons the two stimulus classes are
often referred to as "noise" and "signal", and task performance is
described in terms of hits, correct rejections, omissions, and false
alarms, but this terminology is appropriate only when the model is
applied to tasks that require detection, which is far from always
being the case. In the most widely used version of the model, shown in
Fig.~\ref{basicsdt}, the two evidence distributions are normal with
the same variance, which is usually fixed at unity to make the model
identifiable. The distance $d'$ between the means of the evidence
distributions represents sensitivity. Because normal distributions are
unbounded, $s$ is always ambiguous, and so a criterion $c$ placed on
the evidence axis has to be used to reach a binary decision. A
participant is assumed to decide that a stimulus belongs to the first
class (e.g., an old item) if $s < c$, or that it belongs to the second
class (e.g., a new item) if $s \geq c$. The location of the decision
criterion represents bias.

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot_basicsdt.pdf}
  \caption{Equal-variance normal Signal Detection Theory model}
  \label{basicsdt}
\end{figure}

Perhaps the simplest way of using this model is to fit it to observed
response counts and use the estimated $d'$ values in place of the
percent correct scores; if the model is correct, the resulting
performance measure is not contaminated by bias. Obviously, the model
may not be correct, which is one of the reasons why we focus more on
the generalized version shown in Fig.~\ref{sdtr} below.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot_sdtr.pdf}
  \caption{Equal-variance normal Signal Detection Theory model with
    additional criteria}
  \label{sdtr}
\end{figure}

This generalized model is applicable to studies in which participants
are asked to rate their binary classification decisions on confidence
or some other performance- or stimulus-related dimension. The ratings
and the binary classification decisions can be provided together
(e.g., "I am almost certain that it was a digit"), or in an arbitrary
order.

Ratings are accommodated by introducing additional criteria and
modelling a combined response $y$, which represents both the binary
classification decision and rating. The value of $y$ increases with
the strength of evidence in favor of the second stimulus class. For
example, if confidence is rated on a four-point scale, then $y = 1$
when a subject decides that a stimulus belongs to the first class with
certainty $4$, and $y = 5$ when a subject decides that a stimulus
belongs to the second class with certainty $1$. More formally, if $K$
is the number of possible combined responses, then a subject is
assumed to give response $k$ if $s \in (\bm{c}_{k-1},\bm{c}_k]$, where
$\bm{c}$ are the decision criteria, with $\bm{c}_0$ and $\bm{c}_K$
fixed at $-\infty$ and $+\infty$ respectively.

There is a good reason to collect the ratings and use the generalized
SDT model from Fig.~\ref{sdtr}, even when neither the ratings nor the
placement of criteria are relevant to the research problem. When
$K = 2$ (no ratings), the SDT model fits perfectly, regardless of
whether it is a reasonably good approximation to reality, because the
data and the model have the same dimensionality. This makes the
generalization to the case particularly important, as it is only when
$K > 2$ that the formal assumptions of the model (e.g., equal or
unequal variance) can be tested\footnote{In contrast to the formal
  assumptions, a psychological interpretation of the SDT parameters
  can be tested even when ratings are not available, e.g., by means of
  selective influence \cite{sternberg2001separate}}, which is often
done by comparing observed and predicted ROC curves.

When SDT models are used in psychology studies, researchers are
usually interested in relationships between SDT parameters and
additional measured or manipulated variables; a good example is the
dependence of $d'$ on stimulus strength. Also, in a great majority of
psychology studies in which classification tasks are used, the data
have a hierarchical structure, i.e., there are repeated measures for
each participant or item, and participants or items are only samples
from the target population. It seems worthwhile to explain why a
general-purpose method of inference for SDT models should accommodate
both kinds of situations.

\section{The importance of hierarchical regression structure}

If data have a hierarchical structure, but variability due to
subjects, items, or other sampled factors is not accounted for,
estimates of average (fixed) effects are not guaranteed to be unbiased
and conclusions are not guaranteed to generalize to the target
population. The not uncommon practice of analyzing data aggregated
over sampled factors represents an extreme case of ignoring
hierarchical data structure. The invalidity of this approach in the
context of SDT was clearly illustrated with the results of
simulational studies by \citeA{morey2008problematic}; however,
strictly speaking, it requires such demonstrations only in order to
make a point about a specific case. Because SDT is a non-linear model,
by definition, when estimates of its parameters are based on data
aggregated over sampled factors, the expected values of these
estimates are not in general true population averages. Such estimates
are biased and inference about a target population based on them is
simply not valid. To give a concrete example, consider two unbiased
participants, one with $d'_1 = 2$ and one with $d'_2 = 4$. Their
expected average accuracy is given by
$(\Phi(d'_1/2) + \Phi(d'_2/2)) / 2 = .91$, which corresponds to
$d' = 2.68$, whereas their true average $d'$ is $3$.

Aggregation is not the only way of ignoring a hierarchical data
structure. Sometimes non-aggregated data are analyzed by using
separate estimates for every participant $\times$ item $\times$
condition combination, but uncertainty due to distributions of subject
or item effects is not accounted for by means of a hierarchical model
structure. In such cases, conclusions -- at least with respect to the
uncertainties in estimates of population-average (fixed) effects are
guaranteed to be valid only for the given sample, not the target
population.

Another common source of problems is the separation of estimation of
non-linear model parameters from regression analyses which aim to
relate these parameters to measured or manipulated variables. When the
SDT parameters are estimated separately for each subject and
condition, and only later regressed on predictors of interest, a
number of additional issues may arise.

Firstly, the standard errors or credible intervals associated with the
regression coefficients do not reflect the uncertainty in the SDT
parameter estimates because the latter are treated as mere data
points. The precision of parameter estimates often varies between
participants, items, or conditions, but when the estimates are treated
as data points, no use is made of this information. Secondly,
regressing parameters on numeric predictors makes their estimates
dependent on the common regression structure, and so also on each
other, which can improve the quality of the estimates, just as
assuming that random effects are samples from a common distribution
may improve their estimates.

The aforementioned problems can be dealt with by supplementing the SDT
model with the general hierarchical linear regression structure. For
reasons that we will now explain, this can only be achieved if the
model is substantially reparametrized.

\section{Hierarchical Signal Detection Theory in a constrained
  parameter space}

Both $d'$ and $c$ have the virtue of being directly interpretable in
terms of sensitivity and bias. However, both $d'$ and $c$ are
constrained: $d'$ is non-negative and, when there is more than one
criterion, the elements of the $\bm{c}$ vector are order restricted
($\bm{c}_{i+1} > \bm{c}_i$). Because normal distribution is unbounded,
the hierarchical linear regression structure in which random effects
are normally distributed and fixed effects are represented by
unconstrained parameters can only be defined on unconstrained
parameters. We provide examples of the problems that may otherwise
arise in the following summary of hierarchical SDT implementations.

We are aware of only two attempts at creating a general-purpose method
of inference for hierarchical SDT models with ratings. One is the
Gibbs sampler proposed by \citeA{morey2008problematic} and the other
is the Hierarchical Meta-d' model (HMeta-d') proposed by
\citeA{hmetad}. The HMeta-d' model is a hierarchical version of the
meta-d' model \cite{maniscalco2012signal}, which in turn is a
generalization of the SDT model that allows for a separate
"meta-sensitivity" to account for possible discrepancies between a
binary stimulus classification (referred to as a type 1 task) and the
associated rating task (referred to as a type 2 or meta-cognitive
task). We consider HMeta-d' here because it reduces to the SDT model
with ratings when the type 1 and type 2 sensitivities are equal.

The Gibbs sampler created by \citeA{morey2008problematic} allows for
at most two sampled factors to have independent normally distributed
random effects on the evidence distribution means. Unlike $d'$, each
evidence distribution mean considered in isolation is an unconstrained
parameter, but the mean of the second evidence distribution is by
definition greater than ($d' > 0$) or equal to ($d' = 0$) the mean of
the first. The authors explicitly admit that their algorithm does not
enforce this restriction because, as they claim in the paper, it would
greatly complicate analysis. At the same time the authors fail to
mention that an immediate consequence of allowing for negative $d'$
values is that the resulting posterior distribution loses its intended
interpretation since it can have a non-zero mass when $d' < 0$. The
outermost criteria are fixed at $0$ and $1$, and the ordering
restriction is enforced by assuming that the likelihood is $0$
whenever $\bm{c}_{i+1} \leq \bm{c}_i$. As the authors explain, because
a sampled factor can have independent random effects on the evidence
distribution means, it can have an effect on all the criteria:
shifting both means by the same amount in the same direction is
equivalent to keeping the sensitivity intact, while shifting the
criteria relative to the evidence distributions. However, the elements
of the criteria vector cannot be affected differently by the same
sampled factor, which is an unrealistic restriction.

In HMeta-d' the hierarchical structure is restricted to normally
distributed random intercepts of one sampled factor. Perhaps the most
problematic aspect of the HMeta-d' model, apart from the fact that the
$d'$ parameter is allowed to assume negative values in this model
also, is the representation of the criteria. Each element of the
criteria vector has an associated independent normal distribution,
which does allow for criteria random effects, but does not enforce the
necessary ordering restriction. The elements of this vector are sorted
to obtain another criteria vector, and it is this sorted criteria
vector that is used to model the conditional combined response
distributions. Consequently, the model does contain parameters
representing the actual, order-restricted criteria, but, because
sorting is not injective, the space of the actual criteria is only
loosely related (i.e., not isomorphic) to the space of the
unrestricted criteria vectors that are associated with the
hierarchical structure.

@@

The general hierarchical linear regression structure can be defined on
SDT parameters only if the latter are derived from unconstrained
parameters. In the \code{bhsdtr} package, $d'$ is derived from
$\delta = \ln(d')$, thus random effects on $d'$ can be modelled by
assuming that $\delta$ is normally distributed. The problem of
representing the criteria by unconstrained parameters is solved by
mapping the $R^{K-1}$ space of unconstrained criteria vectors to the
$K$ dimensional probability simplex space using the softmax function,
and mapping the simplex space to the space of order-restricted
criteria vectors by means of the inverse normal CDF:

@@

\noindent Here $i = 1\dots N$ is the observation number, $\bm{X}$ is
the fixed effects model matrix for the respective parameter, $\bm{Z}$
is the random effects model matrix, $\bm{\beta}$ and $\bm{\theta}$ are
the fixed and random effects, $\bm{c}$ is an $N \times K-1$ matrix,
$s$ is the criteria scaling factor, and $\bm{y}$ is the combined
response. Note that $\bm{d'}_i$ is a scalar, but
$\bm{\gamma}_{i,\cdot}$ is in general a vector, and so
$\bm{\beta}^{(\gamma)}$ and $\bm{\theta}^{(\gamma)}$ are matrices. The
$j$-th rows of the $\bm{\beta}^{(\gamma)}$ and
$\bm{\theta}^{(\gamma)}$ matrices represent fixed and random effects
on the $j$-th element of the $\bm{\gamma}$ vector.

Following \citeA{sorensen2015bayesian} we make use of the Cholesky
decomposition of the correlation matrices because it improves
efficiency and admits a convenient prior on random effects
correlations:
